The Warrior Stabs the Goat: Creating a Dothraki Resource Grammar with Grammatical Framework

### 2016-10-14 documentation/paper.tex
Update documentation

\documentclass[a4paper,11pt,twoside]{article}
\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage[british]{babel}

@@

\usepackage{listings}
\lstdefinestyle{customc}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  xleftmargin=\parindent,
  showstringspaces=false,
  basicstyle=\small\ttfamily,
  columns=fullflexible,
  keepspaces=true,
}

@@

@@

\begin{document}

\begin{titlepage}
\clearpage{\pagestyle{empty}\cleardoublepage}
\phantom{bla}
%\vfill
\vfill
\begin{center}
\textsc{\Huge The Warrior Stabs the Goat}
\vspace{0.5cm}

@@

@@

\section{Introduction}
Grammatical Framework (GF) is a functional programming language, 
created in 1998 \cite{gf-book}, doubling as a categorial grammar 
formalism that can be used to implement (natural) language grammars. 
As GF is ``working from a language-independent representation of meaning''
\cite{wiki-gf}
it can be used to generate or parse text in different languages, 
e.g.\ for use in machine translation. 
Since GF uses a symbolic approach to processing and translating language,
in most cases its translations are more precise than those of the 
more commonly used statistical tools \cite{brown-frederking}. 
For example, on somewhat complicated input,
the ``Google Translate'' service \cite{google} will in many cases
produce ungrammatical sentences, while the GF translator 
\cite{gf-translate} has no problems producing 
a correct translation:

\begin{center}
  \begin{tabular}{p{3.7cm}p{4cm}p{3.7cm}}
    Input & Output Google Translate & Output GF Translator\\\hline
    I heal these women. & 
    Ich diese Frauen zu heilen. & 
    Ich verheile diese Frauen.\\
    The warrior will kiss that cheese behind the tree. & 
    Der Krieger wird Kuss, der K\''ase hinter dem Baum. & 
    Der Krieger wird jenen K\''ase hinter dem Baum k\''ussen.\\
    Diese hei\ss e Frau heilt diese K\''onigin im Meer. & 
    This hot woman heals the queen of the sea.  & 
    This hot woman heals this queen in the sea.
  \end{tabular}
\end{center}

@@

\begin{figure}[h]\label{fig-ita}
\centering
  \includegraphics[width=10cm]{ita.png}
  \caption{Mini Resource Grammar Modules and Dependencies \cite[p.\ 209]{gf-book}}
\end{figure}

@@

\begin{figure}[h]\label{fig-rgl}
\centering
  \includegraphics[width=13.5cm]{rgl.jpg}
  \caption{Module Structure of a Full Resource Grammar \cite{gf-slides}}
\end{figure}

@@

  \item Additionally, but optionally, many languages
  also supply a large dictionary of words available in
  that language (allowing application grammar writers
  to use e.g.\ \verb+rise_N+ directly, instead of having to use
  \verb+mk_V ``rise'' ``rose'' ``risen''+ and risking getting
  the irregular forms wrong).
\end{enumerate}

@@

\begin{figure}[h]\label{fig-folders}
\begin{itemize}[itemsep=0.5pt]
\item simpleapi
\item simpleabstract
\item simpledothraki
\item simpleenglish
\item common

@@

\vspace{0.2cm}
\item documentation 
\item vocabulary
\end{itemize}
\caption{The project's folder structure}
\end{figure}

\subsection{Implementation of Grammatical Categories in Dothraki}\label{sec-cats}
In this section we will go into some of the design decisions 
involved in the implementation of the Dothraki resource grammar. 
In order to get a high-level overview, we will first take a look 
at the linearizations of some of the categories defined in the 
API part of the abstract resource grammar library, showing how 
Dothraki grammar can be fit into the GF scheme.

@@

There does not currently seem to be a need to store animacy 
in the agreement record. This is unlike English, where the 
§NP§s gender can affect the verb phrase 
(''The boy loves himself.'' but ``The girl loves herself.'') 
and thus needs to be stored in the §NP§. 
Finally, proper names are (so far as we can tell) 
always animate and third person singular, so none of that 
information needs to be stored in §PN§.

@@

Unlike a §Quant§, a §Det§ has a fixed number.
An interesting feature of determiners in Dothraki is 
that they can appear either before the noun 
(''jinak adra'' = ``this turtle'') or after (such as determiners built 
from possessive pronouns: ``adra anni'' = ``my turtle''). 
Conceivably,\footnote{we cannot currently say for certain, 
due to lack of examples and documentation on 
Dothraki determiner structure} a determiner could even be split 
into two parts (''my two turtles'' $\stackrel{?}=$ ``akat adrasi anni''). 
In anticipation of this, §Det§ has two §Str§ components, §s§ and §s2§. 
Since possessive pronouns are the only examples of postposed 
determiners we know so far, and these happen not to 
inflect with animacy or case, the §s2§ part is simply a §Str§, whereas 
the §s§ part is a table of inflected forms.

@@

\subsubsection{Other categories}

@@

\begin{figure}\label{fig-cats}
  \centering
  \includegraphics[width=13cm]{cats.png}
  \caption{GF Resource Grammar Library Categories \cite{gf-slides}}
\end{figure}

@@

\subsection{Implementation of syntactical functions in Dothraki}

@@

Many of the considerations about Dothraki syntax, 
that have gone into the design of our resource grammar, 
have already been mentioned in \ref{sec-cats}
As one example of how the design of our lincats is used to implement syntactical functions, let us go through the various ways of constructing clauses:

@@

\subsection{Voca

With the current implementation status, it is already possible 
to parse and translate a wide range of phrases. The following 
example text is fully supported by the current version of the 
Dothraki resource grammar:

Beispieltext…

3.  Discussion 
3.1 Limitations
Warum nicht alles auf einmal? Zu viel Arbeit, alles gleichzeitig, keine Möglichkeit zu testen, ….

@@

Incompleteness/overgeneration of GF approach (grammar rules vs. natural language)….
Regeln versuchen so viel wie mögl. abzudecken, aber nicht immer möglich, siehe Buch? Zitat?...
3.2 Outlook
Expand grammar to a full resource grammar. Community?
3.3 Conclusion
Sinnvolles projekt, ambitioniertes ziel, viel gelernt, nicht einfach, stolz…

Appendix
Source Code
The complete source code can be viewed and downloaded via GitHub: 
\url{https://github.com/mahen2/dothraki_gf}

>>>>

\documentclass[a4paper,11pt,twoside]{article}

\makeatletter

\newcommand\frontmatter{%
    \cleardoublepage
  %\@mainmatterfalse
  \pagenumbering{roman}}

\newcommand\mainmatter{%
    \cleardoublepage
 % \@mainmattertrue
  \pagenumbering{arabic}}

\newcommand\backmatter{%
  \if@openright
    \cleardoublepage
  \else
    \clearpage
  \fi
 % \@mainmatterfalse
   }

\makeatother

\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage[british]{babel}

@@

\usepackage{listings}
\lstdefinestyle{customc}{
  belowcaptionskip=1\baselineskip,
  escapeinside=``,
  breaklines=true,
  xleftmargin=\parindent,
  showstringspaces=false,
  basicstyle=\small\ttfamily,
  columns=fullflexible,
  keepspaces=true,
}

@@

\usepackage[table]{xcolor}

\hyphenation{Doth-ra-ki}

@@

\begin{document}
\frontmatter
\begin{titlepage}
\clearpage{\pagestyle{empty}\cleardoublepage}
\phantom{bla}
%\vfill
\begin{center}
\includegraphics[width=5cm]{hhu-logo.jpg}

\vspace{2cm}

{\Large Term Paper}
\vspace{1cm}
\hrule
\vspace{0.4cm}
\textsc{\Huge The Warrior Stabs the Goat}
\vspace{0.5cm}

@@

\vspace{0.4cm}
\hrule
\vspace{2.5cm}

@@

\tableofcontents

\newpage

\thispagestyle{empty}
\begin{center}
\Large
  \includegraphics[width=5cm]{drogo.png}
  \vspace{1.5cm}

  \begin{minipage}{10cm}
  \centering
  ``Khal Drogo shillo chiories she krazaaj. Chiori fin vijazer 
  adraes zoqwe Khales Drogo. Khal Drogo vinde ma chiories ma 
  adraes she krazaaj. Khal Drogo vachrara.''

  \vspace{0.7cm}
  \hrule
  \vspace{0.3cm}

  \color{gray}
``Khal Drogo met a woman on a mountain. The woman, who 
protected a turtle, kissed Khal Drogo. Khal Drogo stabbed the 
woman and the turtle on the mountain. Khal Drogo stinks.''

\vspace{0.5cm}
\color{black}
GF Run-Time System, 2016
  \end{minipage}
\end{center}

\newpage

\thispagestyle{empty}

\mainmatter
\section{Introduction}
Grammatical Framework (GF) is a functional programming language, 
created in 1998 \cite{gf-book}, doubling as a categorial grammar 
formalism that can be used to implement (natural) language grammars. 
As GF is ``working from a language-independent representation of meaning''
\cite{wiki-gf}
it can be used to generate or parse text in different languages, 
e.g.\ for use in machine translation. 
Since GF uses a symbolic approach to processing and translating language,
in most cases its translations are more precise than those of the 
more commonly used statistical tools \cite{brown-frederking}. 
For example, on somewhat complicated input,
the ``Google Translate'' service \cite{google} will in many cases
produce ungrammatical sentences, while the GF translator 
\cite{gf-translate} has no problems producing 
a correct translation:

\begin{center}
  \rowcolors{2}{gray!25}{white}
  \begin{tabular}{p{3.7cm}|p{4cm}|p{3.7cm}}
  \rowcolor{gray!50}
    Input & Output Google Translate & Output GF Translator\\\hline
    I heal these women. & 
    Ich diese Frauen zu heilen. & 
    Ich verheile diese Frauen.\\
    The warrior will kiss that cheese behind the tree. & 
    Der Krieger wird Kuss, der K\"ase hinter dem Baum. & 
    Der Krieger wird jenen K\"ase hinter dem Baum k\"ussen.\\
    Diese hei\ss e Frau heilt diese K\"onigin im Meer. & 
    This hot woman heals the queen of the sea.  & 
    This hot woman heals this queen in the sea.
  \end{tabular}
\end{center}

@@

\begin{figure}[h]
\centering
  \includegraphics[width=10cm]{ita.png}
  \caption{\label{fig-ita}Mini Resource Grammar Modules and Dependencies \cite[p.\ 209]{gf-book}}
\end{figure}

@@

\begin{figure}[h]
\centering
  \includegraphics[width=13.5cm]{rgl.jpg}
  \caption{\label{fig-rgl}Module Structure of a Full Resource Grammar \cite{gf-slides}}
\end{figure}

@@

  \item Additionally, but optionally, many languages
  also supply a large dictionary of words available in
  that language (allowing application grammar writers
  to use e.g.\ \verb+rise_N+ directly, instead of having to use
  \verb+mk_V "rise" "rose" "risen"+ and risking getting
  the irregular forms wrong).
\end{enumerate}

@@

\begin{figure}[h]
\begin{itemize}[itemsep=0.5pt]
\item simpleapi
\item simpleabstract
\item simpledothraki
\item simpleenglish
\item common

@@

\vspace{0.2cm}
\item documentation 
\item vocabulary
\end{itemize}
\caption{\label{fig-folders}The project's folder structure}
\end{figure}

\subsection{Implementation of Grammatical Categories in Dothraki}
\label{sec-impl}
In this section we will go into some of the design decisions 
involved in the implementation of the Dothraki resource grammar. 
In order to get a high-level overview, we will first take a look 
at the linearizations of some of the categories defined in the 
API part of the abstract resource grammar library, showing how 
Dothraki grammar can be fit into the GF scheme.

@@

There does not currently seem to be a need to store animacy 
in the agreement record. This is unlike English, where the 
§NP§s gender can affect the verb phrase 
(``The boy loves himself.'' but ``The girl loves herself.'') 
and thus needs to be stored in the §NP§. 
Finally, proper names are (so far as we can tell) 
always animate and third person singular, so none of that 
information needs to be stored in §PN§.

@@

Unlike a §Quant§, a §Det§ has a fixed number.
An interesting feature of determiners in Dothraki is 
that they can appear either before the noun 
(``jinak adra'' = ``this turtle'') or after (such as determiners built 
from possessive pronouns: ``adra anni'' = ``my turtle''). 
Conceivably,\footnote{we cannot currently say for certain, 
due to lack of examples and documentation on 
Dothraki determiner structure} a determiner could even be split 
into two parts (``my two turtles'' $\stackrel{?}=$ ``akat adrasi anni''). 
In anticipation of this, §Det§ has two §Str§ components, §s§ and §s2§. 
Since possessive pronouns are the only examples of postposed 
determiners we know so far, and these happen not to 
inflect with animacy or case, the §s2§ part is simply a §Str§, whereas 
the §s§ part is a table of inflected forms.

@@

\subsubsection{Other Categories}

@@

\begin{figure}
  \centering
  \includegraphics[width=13cm]{cats.png}
  \caption{\label{fig-cats}GF Resource Grammar Library Categories \cite{gf-slides}}
\end{figure}

@@

\subsection{Implementation of Syntactical Functions in Dothraki}

@@

Many of the considerations about Dothraki syntax, 
that have gone into the design of our resource grammar, 
have already been mentioned in \ref{sec-impl}.
As one example of how the design of our lincats is used to 
implement syntactical functions, let us go through the 
various ways of constructing clauses:

@@

\subsection{Vocabulary Extraction}\label{sec-vocab}

\subsubsection{Dictionaries and Lexicons in the RGL}
As mentioned earlier, many of the resource grammars in GF's
library also come with a dictionary often called \name{Dict\emph{XXX}}.

For example, in \name{DictGer} from the German resource grammar, we
find entries such as:
\begin{lstlisting}
entziehen_V = irregV "entziehen" "entzieht" "entzog" "entz`\"o`ge" 
  "entzogen" ; 
kreis_N = mkN  "Kreis" "Kreise" masculine ;
\end{lstlisting}
An application grammar writer can then use these definitions and does not
have to worry about the gender of nouns, plural forms 
or irregular conjugations.

In fact, the dictionaries in the RGL come in the
form of an abstract grammar with a concrete implementation. The
abstract grammar, often called \name{Dict\emph{XXX}Abs}, contains
entries like
\begin{lstlisting}
entziehen_V : V ;
kreis_N : N ;
\end{lstlisting}

These dictionaries, of course, cannot be used directly for translation.
However, there are two ways in which the RGL supports direct
translation between two langauges. First, there is the abstract
grammar \name{Lexicon} in the \name{abstract} folder, which covers
about 350 more or less commonly used words:
\begin{lstlisting}
...
drink_V2 : V2 ;
dry_A : A ;
dull_A : A ;
dust_N : N ;
...
\end{lstlisting}
This abstract grammar is then implemented in many of the languages
of the RGL. From \name{LexiconGer}:\footnote{The choice of names here is unfortunate:
In the common linguistic sense of the words, \name{DictGer} is a lexicon, while
\name{LexiconGer} is a dictionary...}
\begin{lstlisting}
drink_V2 = dirV2 Irreg.trinken_V ;
dry_A = regA "trocken" ;
dull_A = regA "stumpf" ;
dust_N = reg2N "Staub" "St`\"a`ube" masculine ;
\end{lstlisting}

The \name{Lexicon} module is part of the abstract grammar
\name{Lang} and its concrete implementations
in various languages. Of course, the vocabulary is far
too small to be useful for applications, but it is useful for
experimentation and for testing the resource grammars.

The RGL also has a folder \name{translator} which
mirrors this structure on a much larger scale:
The abstract grammar \name{Dictionary} contains 65,000 English 
words,\footnote{as well as 450 Finnish proper names...} and
the concrete implementations \name{Dictionary\emph{XXX}} contain
their translations into various languages. This is used, e.g.,
in the GF online translator \cite{gf-translate}.

\subsubsection{The Dothraki dictionary}

The publicly available vocabulary list for Dothraki \cite{vocab} currently
contains about 1,500 words.\footnote{This is still not enough to implement
the abstract \name{Lexicon} module from GF's resource grammar library:
Many common words do not yet have a Dothraki equivalent.}
In order to make these accessible for GF application grammars
and for the purpose of translation from and to English, we
supply, as part of our project, a Python script that translates
the wiki source of this list into a GF-compatible format.

The script creates five files in total, adhering closely 
to the structure established in the full RGL, as explained above:
\begin{itemize}
  \item the abstract grammar \name{DictDotAbs.gf} and 
  its concrete implementation \name{DictDot.gf} in the
  folder \name{simpledothraki},
  \item the abstract grammar \name{BigLexicon.gf} in
  \name{simpleabstract} and its implementations
  \name{BigLexiconEng.gf} and \name{BigLexiconDot.gf} in
  \name{simpleenglish} and \name{simpledothraki}, respectively.
\end{itemize}

\begin{figure}
\begin{tabular}{l|l}
\begin{minipage}{.44\linewidth}
\begin{lstlisting}
*'''aranikh''' [a?a'nix]
:''ni.'' subject, topic, theme

*'''gerat''' [ge'?at]
:''vtr.&rarr;abl.'' to lack
:{{Vsup|ger}}
\end{lstlisting}
\end{minipage}&
\begin{minipage}{.55\linewidth}
\begin{lstlisting}
`\textsf{in simpledothraki/DictDotAbs.gf}:`
aranikh_N : N ;
gerat_V2 : V2 ;

`\textsf{in simpledothraki/DictDot.gf}:`
aranikh_N = mkN "aranikh" inanimate ;
gerat_V2 = mkV2 "gerat" "ger" ablative ;

`\textsf{in simpleabstract/BigLexicon.gf}:`
subject_N : N ;
topic_N : N ;
theme_N : N ;
lack_V2 : V2 ;

`\textsf{in simpledothraki/BigLexiconDot.gf}:`
subject_N = DictSimpleEng.subject_N;
topic_N = DictSimpleEng.topic_N;
theme_N = DictSimpleEng.theme_N;
lack_V2 = DictDot.gerat_V2;

`\textsf{in simpleenglish/BigLexiconEng.gf}:`
subject_N = DictDot.aranikh_N;
topic_N = DictDot.aranikh_N;
theme_N = DictDot.aranikh_N;
lack_V2 = DictSimpleEng.lack_V2;
\end{lstlisting}
\end{minipage}
\end{tabular}
\caption{\label{fig-vocab}Two vocabulary wiki entries and generated GF definitions}
\end{figure}

Figure \ref{fig-vocab} shows an example of the GF code the script
generates. A word is added to \name{BigLexicon} whenever a
corresponding word is found in \name{DictEng}, the massive
English dictionary that is part of the RGL.

The wiki vocabulary list has 1,380 entries at the time
of this writing. From these, our script creates
1,410 definitions in \name{DictDot}.\footnote{the somewhat larger
number comes from the fact that many words can be used
in two different ways, for example both as an intransitive
and as a transitive verb, giving more than one entry in \name{DictDot}}

There are a total of 1,916 English translations in the vocabulary list.
The script is able to extract 1,059 entries for \name{BigLexicon}
from this. Many translations cannot be directly added to
\name{BigLexicon} for various reasons:
\begin{itemize}
  \item Some English words appear as translations for
  more than one Dothraki word. The script picks the first
  one it finds.\footnote{Obviously, that might not result in the
  best or most appropriate translation being picked, but that issue
  can be addressed by later manual editing of the generated list.}
  \item Many Dothraki words do not correspond to any one English word.
  For example the list contains entries for ``defensive sword strike'',
  ``sound of fire going out'', ``roast quail'' or
  ``time period of approximately two minutes''. While for some of these,
  it might be desirable to have dictionary 
  entries,\footnote{For example, there could be an entry
  §roast_quail_CN : CN§ in \name{BigLexicon} with linearizations
  §roast_quail_CN = mkCN DictEng.roast_A (mkCN DictEng.quail_N)§
  in
  \name{BigLexiconEng} and §roast_quail_CN = mkCN DictDot.mechikh_N§
  in \name{BigLexiconDot}.} this process is almost impossible
  to automate and must be done by hand.
\end{itemize}

While \name{DictDot} and \name{DictDotAbs} are meant to be used
in application grammars, the \name{BigLexicon\emph{XXX}} files
can be used for testing and translation. An example of this
is shown in the next section.

\subsection{Example}

With the current implementation status, it is already possible 
to parse and translate a wide range of phrases. For experimentation
and testing we supply the \name{LangSimple\emph{XXX}} abstract and
conrete grammars which simply import the \name{SyntaxSimple\emph{XXX}}
and \name{BigLexicon\emph{XXX}} modules (completely
analogous to the \name{Lang\emph{XXX}} module in the
full RGL) and can be used directly
from the GF command line:

\begin{lstlisting}
> i simpleenglish/LangSimpleEng.gf
linking ... OK

Languages: LangSimpleEng
2745 msec
LangSimple> i simpledothraki/LangSimpleDot.gf
linking ... OK

Languages: LangSimpleDot LangSimpleEng
187 msec
LangSimple> p "the warrior which protected the turtle has stabbed my goat with a dagger ." | l
lajak fin viljazer adraes ray vinde dorv anni ma mihesofoon .
the warrior which protected the turtle has stabbed my goat with a dagger .
lajak fin viljazer adraes ray vinde dorv ma mihesofoon anni .
the warrior which protected the turtle has stabbed my goat with a dagger .

31 msec
\end{lstlisting}

The multiple outputs for the last command are 
quite typical and derive from the fact that the parse of the English 
sentence is ambiguous:
Was the duck stabbed with a dagger or was a duck with a dagger stabbed?
Note that, in this case, the Dothraki translations differ, because the
possessive pronoun ``anni'' follows the noun phrase
(either ``dorv'' = ``goat'' or ``dorv ma mihesofoon'' = ``goat with a dagger'').

\section{Discussion}\label{sec-discuss}

To conclude our project documentation, we will now discuss problems 
and limitations as well as briefly describe future plans before 
we summarize the project in its current state and our experience 
while working on it.

\subsection{Limitations}
To begin programming a resource grammar for the first time was 
not easy, since it is not possible to start with, for example, 
just one category and then test the implementation. As there are 
so many dependencies in the resource grammar library, it is recommended 
to take the resource grammar of a similar language and change it step 
by step. But this way, as long as our implementation of the Dothraki 
language was incomplete, the resource grammar would fill the gaps with 
the wrong language grammar. Another way is to start out with an 
independent mini resource grammar, but in that case you cannot make 
use of the resources of the library. Finding our own solution for
this problem took time, but in the end we are satisfied with how
it turned out.

@@

\subsection{Outlook}
In the future, we would like to expand our resource grammar to a
full resource grammar and add it to the GF resource grammar library.
We are already working on and hosting the project online, for everyone
to see, so that the GF and/or Dothraki community may help us to achieve
this goal. Some more parts we can implement right now, for some we have
to wait until more information regarding the Dothraki language is being
published. In the end, we also want to add the full vocabulary to our
dictionary.

\subsection{Conclusion}
Now that the first milestone of our project lies behind us,
we can evaluate the working process and the result. As we had
some starting problems at the beginning of the project, we are
proud to have been able to create such a big portion of the
resource grammar in the given time. By relying on the basics
learned during the seminar and by aiming for such an ambitious
goal, we surely learned a lot during the process. As we are
convinced that our project can be helpful and interesting for
other people in the future, we were motivated to keep on working
on the resource grammar and to try to find the best solution for
every problem. This term paper does not mark the end of this project,
but we are curious to hear what you, our lecturers, and other people
think of the work we have done so far.

\backmatter

\section*{Appendix}

\paragraph{Source Code}

The complete source code can be viewed and downloaded via GitHub: 
\url{https://github.com/mahen2/dothraki_gf}
